---
title: "Group 26: Arcene Cancer"
author: "Todd Davison, Greta Medlam, Douglas Blackwood, Mariam El Mawed"
format:
  html:
    number-sections: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

# Introduction {#sec-Intro}

This file contains code and analysis for the classification of cancerous tissue, based on dimension reduction and various classification methods.

```{r}
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(MASS)
library(smacof)
```


We use the following training and test data sets, that each contain 100 samples of either cancerous or non-cancerous tissue samples.

```{r}
training <- read.csv("group_26.csv")
test <- read.csv("arcene_test.csv")
```


We look at the dimensions of the test and training sets.

```{r}
X_train <- as.matrix(training[,-1])
y_train <- as.matrix(training[,1])

X_test <- as.matrix(test[,-1])
y_test <- as.matrix(test[,1])

data.frame(Training_set = dim(X_train), 
           Test_set = dim(X_test),
           row.names = c("Observations", "Features")) %>%
  gt(rownames_to_stub = T)
```

There are 2506 features and 100 observations in the training set. We look at the split of cancerous to non-cancerous observations.

```{r}
cancerous <- training[training$Class == 1,]
dim(cancerous)
```

So we have 44 cancerous tissue samples and 56 non-cancerous samples.

We will now check if the variables are on the same scale.

```{r}
# standard deviations of variables
sds <- apply(X_train, 2, sd, na.rm = TRUE)
summary(sds)
```

There is a large difference between the min and max, so we look at a boxplot.

```{r}
boxplot(sds, main = "SDs of variables")
```

From this we can see the variables have different scales, so we will use the correlation matrix instead of covariance matrix for PCA and we will scale the data when using MDS.

## Dimension Reduction

We perform PCA using the correlation matrix.

```{r}
#| warning: false
#| message: false
pca_fit <- prcomp(X_train, cor = T)
pca_summary <- summary(pca_fit)
pca_summary$importance[, 1:20]
```

From this we can see the first 6 principal components (PCs) explain 80% of the variability, and 18 explains 90% of the variability. 

We will now perform Classical MDS.

```{r}
# Euclidean distances for training points
D_train <- dist(scale(X_train), method = "euclidean")

# classical MDS to k dimensions
cmds_fit <- cmdscale(D_train, k = 10)
summary(cmds_fit)
```

To evaluate the information retained we look at the correlation between our new point and the original points.

```{r}
orig_dist <- as.vector(D_train)
mds_dist <- as.vector(dist(cmds_fit))

cor(orig_dist, mds_dist)
```

There is very strong correlation between the distance between the original points and the new lower dimensional points. The correlation can be made stronger by increasing k.

Metric MDS:

```{r}
corr_mat <- cor(X_train)
dissimilarity <- sim2diss(corr_mat, method = "corr")
mmds_fit_2 <- mds(dissimilarity, ndim = 2, type = "interval")
mmds_fit_5 <- mds(dissimilarity, ndim = 5, type = "interval")
mmds_fit_10 <- mds(dissimilarity, ndim = 10, type = "interval")
mmds_fit_15 <- mds(dissimilarity, ndim = 15, type = "interval")

data.frame(ndim = c(2,5,10,15),stress_score = c(mmds_fit_2$stress, mmds_fit_5$stress, mmds_fit_10$stress, mmds_fit_15$stress)) %>% gt()
```

We can see the difference between 10 and 15 dimensions doesn't reduce the cost function by much, so a sensible choice for ndim likely lies in this region. Below are some Shepard plots.

```{r}
par(mfrow = c(2, 2))   # 2 rows, 2 columns
plot(mmds_fit_2,  plot.type="Shepard", main="Shepard diag. 2D")
plot(mmds_fit_5,  plot.type="Shepard", main="Shepard diag. 5D")
plot(mmds_fit_10,  plot.type="Shepard", main="Shepard diag. 10D")
plot(mmds_fit_15,  plot.type="Shepard", main="Shepard diag. 15D")
```

Ideally these should be a straight line. We see 15 dimensions agrees with this the most. 