---
title: "Group 26: Arcene Cancer"
author: "Todd Davison, Greta Medlam, Douglas Blackwood, Mariam El Mawed"
format:
  html:
    number-sections: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

# Introduction {#sec-Intro}

This file contains code and analysis for the classification of cancerous tissue, based on dimension reduction and various classification methods.

```{r}
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(MASS)
library(smacof)
library(ROCR)
```

We use the following training and test data sets, that each contain 100 samples of either cancerous or non-cancerous tissue samples.

```{r}
training <- read.csv("group_26.csv")
test <- read.csv("arcene_test.csv")
```

We look at the dimensions of the test and training sets.

```{r}
X_train <- as.matrix(training[,-1])
y_train <- as.matrix(training[,1])

X_test <- as.matrix(test[,-1])
y_test <- as.matrix(test[,1])

data.frame(Training_set = dim(X_train), 
           Test_set = dim(X_test),
           row.names = c("Observations", "Features")) %>%
  gt(rownames_to_stub = T)
```

There are 2506 features and 100 observations in the training set. We look at the split of cancerous to non-cancerous observations.

```{r}
cancerous <- training[training$Class == 1,]
dim(cancerous)
```

So we have 44 cancerous tissue samples and 56 non-cancerous samples.

We will now check if the variables are on the same scale.

```{r}
# standard deviations of variables
sds <- apply(X_train, 2, sd, na.rm = TRUE)
summary(sds)
```

There is a large difference between the min and max, so we look at a boxplot.

```{r}
boxplot(sds, main = "SDs of variables")
```

From this we can see the variables have different scales, so we will use the correlation matrix instead of covariance matrix for PCA and we will scale the data when using MDS.

## Dimension Reduction

We perform PCA using the correlation matrix.

```{r}
#| warning: false
#| message: false
pca_fit <- prcomp(X_train, scale. = T)
pca_summary <- summary(pca_fit)
pca_summary$importance[, 1:50]
```

From this we can see the first 40 principal components (PCs) explain 80% of the variability, and 50 explains 85% of the variability.

Since 40 PCs explain 80% of the variability, we will proceed with those ones.

```{r}
#| warning: false
#| message: false
#Select/Choose the first 18 pcs which explains 90% of the variation

k <- 40   # number of PCs you decided to keep

X_train_pca <- pca_fit$x[, 1:k]

# Test PCA scores
X_test_pca <- predict(pca_fit, newdata = X_test)[, 1:k]

```

Now this new X_train_pca and \_train will be used to train the classification models and test them using X_test_pca and y_test.

## Support Vector Machines

Initially we will begin with SVM with 18 principal components, default cost, linear kernel and C-classification.

```{r}
# Initially we use 18 pc's
n_pcs   <- 18
train_data <- as.data.frame(pca_fit$x[, 1:n_pcs])

train_data$Class <- y_train

test_data  <- as.data.frame(predict(pca_fit, X_test)[, 1:n_pcs])
test_data$Class <- y_test

# Call SVM
svm_18pc <- svm(Class ~ ., data = train_data, type = "C-classification", kernel = "linear")
svm_18pc
```

Now we test the model, using confidence scores, ROC and AUC. A good model that predicts the cancerous tumors well will have high AUC.

```{r}
SVM_model1 <- svm(Class ~ ., data = train_data, type = "C-classification", kernel = "linear", cost = 10, decision.values = TRUE)

pred_class_model1 <- predict(SVM_model1, test_data)
conf_mat_model1   <- table(Actual = test_data$Class, Predicted = pred_class_model1)
accuracy_model1   <- sum(diag(conf_mat_model1)) / sum(conf_mat_model1) * 100


# Get scores on the TEST data on AUC
scores_model1 <- attributes(predict(SVM_model1, test_data, decision.values = TRUE))$decision.values

# ROC plot and AUC calulation
# Use ROCR:: prefix to force R to use the right functions
pred_model1 <- ROCR::prediction(scores_model1, test_data$Class)
perf_model1 <- ROCR::performance(pred_model1, "tpr", "fpr")
plot(perf_model1, main = "ROC Curve")

auc_model1 <- ROCR::performance(pred_model1, "auc")@y.values[[1]]


auc_model1
accuracy_model1
```

For model 1, which we will use as our baseline model, we see an AUC of \$\\approx\$ 0.89. This is a well fitting model to start off with.

We will now try to optimise the model through comparison of cost, linear vs non-linear boundary, class imbalance and principal components.

Class Imbalance assessment:

```{r}
set.seed(2003)
weighted.tune <- tune(svm, Class ~., data = train_data,
                      type = "C-classification", kernel = "linear",
                      ranges = list(cost = c(0.1, 1, 10, 100)),
                      class.weights = c("-1" = 1, "1" = 2))
summary(weighted.tune)

```

We will try to compare non-linear model vs linear model

```{r}
train_data$Class <- as.factor(train_data$Class)
test_data$Class  <- as.factor(test_data$Class)
radial.tune <- tune(svm, Class ~., data = train_data,
                    type = "C-classification", kernel = "radial",
                    ranges = list(cost  = c(0.1, 1, 10, 100),
                                  gamma = c(0.5, 1, 2, 3, 4)))

summary(radial.tune)

# We see minimum error at cost = 10, gamma = 0.5 and cost = 100, gamma = 0.5
# We will test both of these cases against the linear model

m_radial_c10 <- svm(Class ~ ., data = train_data, type = "C-classification",
                kernel = "radial", cost = 10, gamma = 0.5,
                decision.values = TRUE)

s_c10<- attributes(predict(m_radial_c10, test_data, decision.values = TRUE))$decision.values
p_c10 <- ROCR::prediction(s_c10, test_data$Class)
auc_radial_c10 <- ROCR::performance(p_c10, "auc")@y.values[[1]]
auc_radial_c10

# c = 100

m_radial_c100 <- svm(Class ~ ., data = train_data, type = "C-classification",
                kernel = "radial", cost = 100, gamma = 0.5,
                decision.values = TRUE)

s_c100<- attributes(predict(m_radial_c100, test_data, decision.values = TRUE))$decision.values
p_c100 <- ROCR::prediction(s_c100, test_data$Class)
auc_radial_c100 <- ROCR::performance(p_c100, "auc")@y.values[[1]]
auc_radial_c100
```

We get a slightly higher value of AUC in the linear boundary model, hence we will proceed with this model.

Cost optimisation:

```{r}
# Optimising Cost ---------------------------------------------------------

# To use tune we need the class variables as factors in both data sets

set.seed(2003)
train_data$Class <- as.factor(train_data$Class)
test_data$Class  <- as.factor(test_data$Class)



linear.tune<-tune(svm, Class~., data=train_data,
                  type="C-classification", kernel="linear",
                  ranges=list(cost=c(0.1,1,10,100)))
summary(linear.tune)

# We see from this that the best cost value will be 0.1, as error is lowest at 0.18, also dispersion is lowest, we notice weeor increasing monotomicaly so we may try smaller cost to see if trend continues.

linear.tune2 <-tune(svm, Class~., data=train_data,
                  type="C-classification", kernel="linear",
                  ranges=list(cost=c(0.01, 0.03, 0.05, 0.07, 0.1)))
summary(linear.tune2)

# We see that for this somewhere between cost = 0.05 and 0.1 gives us the lowest error, We will try again within this range.


linear.tune3 <-tune(svm, Class~., data=train_data,
                  type="C-classification", kernel="linear",
                  ranges=list(cost=c(0.06, 0.07, 0.08, 0.09)))
summary(linear.tune3)

# Note: 0.07 -> 0.1 all have same error, we wioll select cost then on lowest dispersion. Hence cost = 0.1

# Fitting new model to compare to model 1
SVM_model2 <- svm(Class ~ ., data = train_data, type = "C-classification",
                kernel = "linear", cost = 0.1, decision.values = TRUE)

scores_model2 <- attributes(predict(SVM_model2, test_data, decision.values = TRUE))$decision.values

pred_model2 <- ROCR::prediction(scores_model2, test_data$Class)
perf_model2 <- ROCR::performance(pred_model2, "tpr", "fpr")
plot(perf_model2, main = "ROC Curve (cost = 0.1)")
auc_model2 <- ROCR::performance(pred_model2, "auc")@y.values[[1]]
auc_model2
```

Through optimising cost, we see a slight increase in the AUC value, suggesting a more accurate model.
