---
title: "Group 26: Arcene Cancer"
author: "Todd Davison, Greta Medlam, Douglas Blackwood, Mariam El Mawed"
format:
  html:
    number-sections: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

# Introduction {#sec-Intro}

This file contains code and analysis for the classification of cancerous tissue, based on dimension reduction and various classification methods.

```{r}
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(MASS)
library(smacof)
library(ROCR)
library(neuralnet)
library(e1071)
library(rpart)
library(rpart.plot)
library(randomForest)
```

We use the following training and test data sets, that each contain 100 samples of either cancerous or non-cancerous tissue samples.

```{r}
training <- read.csv("group_26.csv")
test <- read.csv("arcene_test.csv")
```

We look at the dimensions of the test and training sets.

```{r}
X_train <- as.matrix(training[,-1])
y_train <- as.matrix(training[,1])

X_test <- as.matrix(test[,-1])
y_test <- as.matrix(test[,1])

data.frame(Training_set = dim(X_train), 
           Test_set = dim(X_test),
           row.names = c("Observations", "Features")) %>%
  gt(rownames_to_stub = T)
```

There are 2506 features and 100 observations in the training set. 

We now look for missing data.

```{r}
X_train_df <- as.data.frame(X_train)

total_missing <- sum(is.na(X_train_df))
```

No missing data.

We look at the split of cancerous to non-cancerous observations.

```{r}
cancerous <- training[training$Class == 1,]
dim(cancerous)
```

So we have 44 cancerous tissue samples and 56 non-cancerous samples.

We also look at the test set split.

```{r}
cancerous <- test[test$Class == 1,]
dim(cancerous)
```

We look at the correlation between variables, as PCA works best when variables are highly correlated.


```{r}
corr_mat <- cor(X_train)
cors <- corr_mat[upper.tri(corr_mat)]

mean_cor <- mean(cors)
mean_cor
```

This is very low. However there are dummy variables in the dataset, so this mean may be skewed by dummy variables with very little/no correlation to non-dummy variables.

```{r}
high_count <- sum(abs(cors) >= 0.67)
mid_count <- sum(abs(cors) >= 0.33 & abs(cors) < 0.67)
little_count <- sum(abs(cors) >= 0.01 & abs(cors) < 0.33)
zero_count   <- sum(abs(cors) < 0.01)

c(
  Highly_correlated = high_count,
  Mild_correlation = mid_count,
  Little_correlation = little_count,
  Near_zero_correlation = zero_count
) 
```

While a majority of variables have little to no correlation there is a significant amount with high or mild correlation, so PCA should still work reasonably well.

We will now check if the variables are on the same scale.

```{r}
# standard deviations of variables
sds <- apply(X_train, 2, sd, na.rm = TRUE)
summary(sds)
```

There is a large difference between the min and max, so we look at a boxplot.

```{r}
boxplot(sds, main = "SDs of variables")
```

From this we can see the variables have different scales, so we will use the correlation matrix instead of covariance matrix for PCA and we will scale the data when using MDS.

## Dimension Reduction

We perform PCA using the correlation matrix.

```{r}
#| warning: false
#| message: false
pca_fit <- prcomp(X_train, scale = T)
pca_summary <- summary(pca_fit)
pca_summary$importance[, 1:50]
```

From this we can see the first 40 principal components (PCs) explain 80% of the variability, and 50 explains 85% of the variability.

We look at a scree plot.

```{r}
# Proportion of variance explained
pve <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2)

# Scree plot
plot(
  1:50, pve[1:50],
  xlab = "Principal Component",
  ylab = "Proportion of Variance Explained",
  main = "Scree Plot"
)
```

From the scree plot we can see the "elbow" appears between 8 and 15 PCs approximately.

Since 40 PCs explain 80% of the variability, we will proceed with this as our default number of PCs.

```{r}
#| warning: false
#| message: false
#Select/Choose the first 40 pcs which explains 90% of the variation

k <- 40   # number of PCs you decided to keep

X_train_pca <- pca_fit$x[, 1:k]

# Test PCA scores
X_test_pca <- predict(pca_fit, newdata = X_test)[, 1:k]

```

We can also use MDS for dimension reduction, however it doesn't have a predict function to map data from the test set to the same space that the model was trained on. Hence, we will not use it for classification models.

We will now perform Classical MDS.

```{r}
# Euclidean distances for training points
D_train <- dist(scale(X_train), method = "euclidean")

# classical MDS to k dimensions
cmds_fit <- cmdscale(D_train, k = 90, eig = T)
cmds_fit$GOF
```

This gives a very good goodness of fit value for k = 90.

We will now look at various classification methods, using PCA.

## Support Vector Machines

Support Vector Machines (SVM) is a supervised learning technique used to classify data. Here we will be using SVM to classify the variable 'Class'. Class is conveniently in the format already required by SVM of being binary with values -1, 1 for non-cancerous and cancerous respectively.

In the following code, different kernels will be tuned based on error and compared using 10-fold cross validation. Each SVM model will have AUC calculated based on ROC plot, as a measure of 'goodness of fit', as such models will be compared. Through comparison of confusion matrix calculated for each model, the model selection/optimisation process can take place to determine the best SVM classification model to detect cancerous cells in the test set.

Initially, a baseline AUC value must be caluated using default cost, C-classification and linear kernel.

```{r}
# Preprocess data for SVM and set seed
# Initially we use 40 pc's

n_pcs <- 40
train_data <- as.data.frame(pca_fit$x[, 1:n_pcs])

train_data$Class <- y_train

test_data  <- as.data.frame(predict(pca_fit, X_test)[, 1:n_pcs])

# SVM must have repsonse as factors
test_data$Class <- as.factor(y_test)
```

Fit Baseline SVM:

```{r}
set.seed(1)
# test baseline model, plotting ROC and calculating AUC
SVM_model1 <- svm(Class ~ ., data = train_data, type = "C-classification", kernel = "linear", cost = 1, decision.values = TRUE)

# Calcualte confidence scores for ROC curve
scores_model1 <- attributes(predict(SVM_model1, test_data, decision.values = TRUE))$decision.values


# Use ROCR:: 'ROCR::' as had error woth libraries
pred_model1 <- ROCR::prediction(scores_model1, test_data$Class)
perf_model1 <- ROCR::performance(pred_model1, "tpr", "fpr")

# Plot Roc curve for Baseline Model
par(mfrow = c(1, 1))
plot(perf_model1, main = "ROC Curve (Baseline)")

# Calcualte AUC for baseline model
auc_model1 <- ROCR::performance(pred_model1, "auc")@y.values[[1]]

auc_model1
# 0.88


# Classification error and confusion matrix
pred_class_model1 <- predict(SVM_model1, test_data)
error_model1 <- mean(pred_class_model1 != test_data$Class)
error_model1
table(ActualClass = test_data$Class, Prediction = pred_class_model1)

# We get 29 TP, 50 TN, 6 FP, 15 FN 
```

For model 1, our baseline model, we see an AUC of approximately 0.88. This is a well fitting model to start off with. We see that this model has 15 false negative values.

We will now try to optimise the model through comparison of cost, linear vs non-linear boundary and principal components.

Linear Cost Analysis: Cost value tells the function how many miss classifications are allowed i.e. false positive or false negative. Higher cost value can result in overfitting. We want to optimise cost so we can reduce the false negative results.

```{r}
# To use tune we need the class variables as factors in both data sets

train_data$Class <- as.factor(train_data$Class)
test_data$Class  <- as.factor(test_data$Class)


# use tune function for cross-validation testing of different cost values
set.seed(1)
linear.tune<-tune(svm, Class~., data=train_data,
                  type="C-classification", kernel="linear",
                  ranges=list(cost=c(0.1,1,10,100)))
summary(linear.tune)
summary(linear.tune)$best.parameters

# We see from this that the best cost value will be 0.1, as error is lowest at 0.26, also dispersion is lowest, we notice however, error has a positive trend with cost so we may try smaller cost to see if trend continues.

set.seed(1)
linear.tune2 <-tune(svm, Class~., data=train_data,
                  type="C-classification", kernel="linear",
                  ranges=list(cost=c(0.01, 0.03, 0.05, 0.07, 0.1)))
summary(linear.tune2)
summary(linear.tune2)$best.parameters

# We see that for this somewhere between (not including) cost = 0.01 to cost = 0.07 gives us the lowest error, We will try again within this range.

set.seed(1)
linear.tune3 <-tune(svm, Class~., data=train_data,
                  type="C-classification", kernel="linear",
                  ranges=list(cost=c(0.02, 0.03, 0.04, 0.05, 0.06)))
summary(linear.tune3)

# We see that error occurs at cost = 0.03, 0.05, 0.06, with the latter having the lowest dispersion, hence cost = 0.06 will be the value going forward.

# Fitting new cost tuned model to compare to model 1
set.seed(1)
SVM_model2 <- svm(Class ~ ., data = train_data, type = "C-classification",
                kernel = "linear", cost = 0.06, decision.values = TRUE)

scores_model2 <- attributes(predict(SVM_model2, test_data, decision.values = TRUE))$decision.values

pred_model2 <- ROCR::prediction(scores_model2, test_data$Class)
perf_model2 <- ROCR::performance(pred_model2, "tpr", "fpr")

par(mfrow = c(1, 1))
plot(perf_model2, main = "ROC Curve (cost = 0.06)")
auc_model2 <- ROCR::performance(pred_model2, "auc")@y.values[[1]]
auc_model2
# 0.898
# Model 2 better than model 1


# Classification error and confusion matrix
pred_class_linear <- predict(SVM_model2, test_data)
error_linear <- mean(pred_class_linear != test_data$Class)
error_linear
table(ActualClass = test_data$Class, Prediction = pred_class_linear)
```

Here we see a slight improvement of AUC when compared the the baseline model. It is also worth noting that the false negative count here is 10, which is an impovement on the baseline model meaning we miss less cancers with this model.

Radial Kernel Analysis: This allows testing of different boundary types. Again, the kernel will be assessed on both AUC and false negative rate, with the latter being more influential in the optimisation of the model.

```{r}
set.seed(1)
train_data$Class <- as.factor(train_data$Class)
test_data$Class  <- as.factor(test_data$Class)

radial.tune <- tune(svm, Class ~., data = train_data,
                    type = "C-classification", kernel = "radial",
                    ranges = list(cost  = c(0.1, 1, 10, 100),
                                  gamma = c(0.001, 0.01, 0.05, 0.1, 0.5)))
summary(radial.tune)


# We see minimum error at cost = 1, gamma = 0.05
# We will test both of these cases against the linear model
set.seed(1)
svm_radial <- svm(Class ~ ., data = train_data, type = "C-classification",
                kernel = "radial", cost = 1, gamma = 0.05,
                decision.values = TRUE)

scores_radial <- attributes(predict(svm_radial, test_data, decision.values = TRUE))$decision.values

pred_radial <- ROCR::prediction(scores_radial, test_data$Class)
perf_radial <- ROCR::performance(pred_radial, "tpr", "fpr")
par(mfrow = c(1, 1))
plot(perf_radial, main = "ROC Curve (Radial)")
auc_radial <- ROCR::performance(pred_radial, "auc")@y.values[[1]]
auc_radial
# 0.908
# This is better than the linear model based on AUC

# Classification error and confusion matrix
pred_class_radial <- predict(svm_radial, test_data)
error_radial <- mean(pred_class_radial != test_data$Class)
error_radial
table(ActualClass = test_data$Class, Prediction = pred_class_radial)
# We get very few false negatives: this is good for cancer
```

The raidal RBF kernel, has a greater AUC than the tuned linear model, along with producing only 3 false negative counts, this means that of the 100 tumours, we only miss three cancerous ones, while correctly identifying 41 cancers. There are 19 false positive rates, but given the context of the problem this is not as alarming as it may seem as the most dangerous inaccuracy is false negative.

Degree 2 Polynomial Kernel Test: This is similar to the above test but for polynomial of degree 2

```{r}
# Tune polynomial kernel (degree 2) over cost, gamma and coef0
set.seed(1)
poly.tune <- tune(svm, Class ~., data = train_data,
                  type = "C-classification", kernel = "polynomial", degree = 2, ranges = list(cost  = c(0.1, 1, 5), 
                                                                                              gamma = c(0.01, 0.05, 0.1), 
                                                                                              coef0 = c(0, 1, 2, 3)))
summary(poly.tune)
poly.tune$best.model

# creat updated SVM:
svm_poly <- svm(Class ~ ., data = train_data, type = "C-classification", kernel = "polynomial",degree = 2, cost = 1, 
                  gamma = 0.05, coef0 = 2, decision.values = TRUE)

# Evaluate on test set
scores_poly <- attributes(predict(svm_poly, test_data, decision.values = TRUE))$decision.values
pred_poly <- ROCR::prediction(scores_poly, test_data$Class)
perf_poly <- ROCR::performance(pred_poly, "tpr", "fpr")

par(mfrow = c(1, 1))
plot(perf_poly, main = "ROC Curve (Deg. 2 Polynomial)")

auc_poly <- ROCR::performance(pred_poly, "auc")@y.values[[1]]
auc_poly
# 0.886

# Classification error and confusion matrix
pred_class_poly <- predict(svm_poly, test_data)
error_poly <- mean(pred_class_poly != test_data$Class)
error_poly
table(ActualClass = test_data$Class, Prediction = pred_class_poly)
```

The AUC value is very slightly lower for the polynomial kernel SVM than for the radial kernel, suggesting radial is more accurate. Furthermore, there are 11 false negatives, over three times that of the radial kernel model.

Thus, the best model is the radial model.

## Neural Network

We will now use a Neural Network (NN) to predict a classification. We will use a validation set to decide on the NN parameters, i.e. PCs used and number of hidden nodes. We use an 80/20 split.

```{r}
set.seed(123)

k <- 80

# 80/20 split
index <- sample(1:nrow(X_train), round(0.8 * nrow(X_train)))

# New test and valiation sets
X_subtrain <- X_train[index,]
X_valid    <- X_train[-index,]

# Classifications for new sets
y_subtrain <- y_train[index]
y_valid    <- y_train[-index]

# Remove zero-variance columns (causes an error when running PCA if we don't)
subtrain_sd <- apply(X_subtrain, 2, sd)
keep_cols <- subtrain_sd > 0

X_subtrain <- X_subtrain[, keep_cols]
X_valid    <- X_valid[, keep_cols]

# PCA on sub-training set 
pca_fit <- prcomp(X_subtrain, scale. = TRUE)

# Define the Principal components for training and validation sets
train_pcs <- pca_fit$x[, 1:k]
valid_pcs <- predict(pca_fit, newdata = X_valid)[, 1:k]
```

We change the classification from -1 or 1, which it is in the original data, to 0 or 1 as this is what neuralnet assumes. We then construct the training and test sets for the NN.

```{r}
y_subtrain_bin <- ifelse(y_subtrain == "-1", 0, 1)
y_valid_bin  <- ifelse(y_valid  == "-1", 0, 1)

train_nn <- data.frame(Class = y_subtrain_bin, train_pcs)
valid_nn  <- data.frame(Class = y_valid_bin,  valid_pcs)
```

We ensure the training and validation sets contain observations of both classes.

```{r}
mean(y_subtrain_bin == 0)
mean(y_valid_bin == 0 )
```

Both have a roughly equal split so we continue.

We fit an initial NN using the logistic function and cross entropy loss.

```{r}
set.seed(123)
# Fit neural network
nn_fit <- neuralnet(
  Class ~ .,
  data = train_nn,
  hidden = c(5),          # 1 hidden layer with 5 neurons
  linear.output = FALSE,  # classification
  act.fct = "logistic",
  err.fct = "ce",         # cross-entropy
  likelihood = T
)
CE_loss <- round(nn_fit$result.matrix[1,1],3)
AIC <- round(nn_fit$result.matrix[4,1],3)
BIC <- round(nn_fit$result.matrix[5,1],3)

tibble('Metric' = c('AIC', 'BIC','CE Loss'),
                       'Value' = c(AIC, BIC, CE_loss)) %>% gt()
```

We will now look at the effect of varying the number of PCs used and the number of hidden nodes.

```{r}
set.seed(123)
results <- data.frame()

for (k in 10:80) {
 # training PCs
  train_pcs <- pca_fit$x[, 1:k]
  
  # validation PCs projected into the same PCA space
  valid_pcs <- predict(pca_fit, newdata = X_valid)[, 1:k]
  
  # data frames for neuralnet
  train_nn <- data.frame(Class = y_subtrain_bin, train_pcs)
  valid_nn <- data.frame(Class = y_valid_bin, valid_pcs)
  
  for (h in 5:15) {
    
    nn_fit <- neuralnet(
      Class ~ .,
      data = train_nn,
      hidden = c(h),
      linear.output = FALSE,
      act.fct = "logistic",
      err.fct = "ce",
      likelihood = TRUE
    )
    
    # numerical summaries
    CE_loss_train <- nn_fit$result.matrix[1, 1]
    AIC_train <- nn_fit$result.matrix[4, 1]
    BIC_train <- nn_fit$result.matrix[5, 1]
    
    # validation predictions
    valid_prob <- as.numeric(predict(nn_fit, valid_nn[,-1]))
    
    # convert probabilities into classification (i.e. assign to case if probability > 50%)
    valid_pred <- ifelse(valid_prob >= 0.5, 1, 0)
    
    valid_accuracy <- mean(valid_pred == y_valid_bin)
    
    results <- rbind(
      results,
      data.frame(
        PCs = k,
        hidden = h,
        AIC = AIC_train,
        BIC = BIC_train,
        TrainCrossEntropyLoss = CE_loss_train,
        ValidationAccuracy = valid_accuracy
      )
    )
  }
}

results %>%
  arrange(desc(ValidationAccuracy), hidden) %>%
  slice_head(n = 10) %>%
  gt()
```

Since multiple NNs get a validation accuracy score of 90% we use the simplest one, with 17 PCs and 11 hidden nodes. We define simplest here as the lowest AIC and BIC score. Note the validation accuracy changes by 0.05 as we only have 20 observations in the validation set, hence identifying one more observation correctly changes the score by 1 / 20 = 0.05. Now we fit our model.

```{r}
set.seed(123)

# Number of PCs
k <- 17
# Number of hidden nodes
h <- 11

# Perform PCA on whole training set
pca_fit <- prcomp(X_train, scale. = T)

# Define the Principal components for training and validation sets
train_pcs <- pca_fit$x[, 1:k]
test_pcs  <- predict(pca_fit, newdata = X_test)[, 1:k]

# Update classification to 0 or 1 (as required by neuralnet)
y_train_bin <- ifelse(y_train == "-1", 0, 1)
y_test_bin  <- ifelse(y_test  == "-1", 0, 1)

train_nn <- data.frame(Class = y_train_bin, train_pcs)
test_nn  <- data.frame(Class = y_test_bin, test_pcs)

# Create final neural network with optimised parameters
final_nn <- neuralnet(
  Class ~ .,
  data = train_nn,
  hidden = c(h),
  linear.output = FALSE,
  act.fct = "logistic",
  err.fct = "ce",
  likelihood = TRUE
)

# Make classifications on test set based on our NN
test_prob <- as.numeric(predict(final_nn, test_nn[,-1]))
test_pred <- ifelse(test_prob >= 0.5, 1, 0)

# Compare our predicted classifications to actual
test_accuracy <- mean(test_pred == y_test_bin)
test_accuracy

cm <- table(Predicted = test_pred, Actual = y_test_bin)
cm
```

This gives a test accuracy of 87%. We will now look at the false positives/negatives rate.

```{r}
TN <- cm["0", "0"]
FN <- cm["0", "1"]
FP <- cm["1", "0"]
TP <- cm["1", "1"]

false_positive_rate <- FP / (FP + TN)
false_negative_rate <- FN / (FN + TP)

false_positive_rate
false_negative_rate
```

This gives a false negative rate of roughly 0.14.

## Classification Trees

We now use binary classification trees to predict the class of tissue samples. To begin, trees will be fit to the data for different numbers of PCS. A topdown approach will be taken to fit full trees and prune them according to the minumum error criterion, which produces more accurate results than the smallest tree criterion.

First we use 40 PCs, which capture 80% of variability in the data

```{r}
#Apply PCA to our training and test data
PCA40_train <- data.frame(Class = as.factor(y_train), predict(pca_fit, X_train)[,1:40])
PCA40_test <- data.frame(Class = as.factor(y_test), predict(pca_fit, X_test)[,1:40])

# Full tree:
tree_PCA40_full <- rpart(Class~., data=PCA40_train, method="class", minsplit=2, 
                        minbucket=1, cp=-1)
tree_PCA40_full

# Pruning:
printcp(tree_PCA40_full)
# Minimum xerror: Tree 6, cp=0.022727
tree_PCA40 <- prune(tree_PCA40_full, cp=0.023)
rpart.plot(tree_PCA40, type=2, extra=4 )
tree_PCA40
```

Assessing variable importance in this tree shows that PC13 is the most important variable (13.75), with most PCs beyond 13 holding very little importance. PC38 is the only important PC above 13 (6.46).

```{r}
tree_PCA40$variable.importance 
# PC13 is significantly most important, 13 PCs only explain 62% of variability though
# PC38 is relatively important, but maybe not enough to justify that many more?
```

We now assess the accuracy of this tree by using it to predict the class of samples in the test data. The test accuracy of this tree is 69%, not a particularly good score.

```{r}
# Now to assess overall accuracy of this tree
tree_PCA40_pred <- predict(tree_PCA40, newdata=PCA40_test, type="class")
class_rate_tree40 <- mean(tree_PCA40_pred==PCA40_test$Class)
class_rate_tree40 
```

Before fitting different tree based methods to improve this accuracy, it is worth investigating whether using a different number of PCs improves accuracy.

We repeat the process using 50 PCs, which capture 85% of the variability in the data.

```{r}
PCA50_train <- data.frame(Class = as.factor(y_train), predict(pca_fit, X_train)[,1:50])
PCA50_test <- data.frame(Class = as.factor(y_test), predict(pca_fit, X_test)[,1:50])

# Full tree:
tree_PCA50_full <- rpart(Class~., data=PCA50_train, method="class", minsplit=2, 
                         minbucket=1, cp=-1)
tree_PCA50_full

# Pruning:
printcp(tree_PCA50_full)
# Minimum xerror: Tree 6, cp=0.022727
# There appears to be an optimal complexity parameter
tree_PCA50 <- prune(tree_PCA50_full, cp=0.023)
rpart.plot(tree_PCA50, type=2, extra=4 )
tree_PCA50
```

We see that minimum error pruning results in the same complexity parameter as before. We also see *identical* variable importance as the previous tree

```{r}
tree_PCA50$variable.importance
# Similar to previous tree
# 13 at the top
```

And the test accuracy is 69% again, so this level of PCs leads to no improvement

```{r}
# Overall accuracy
tree_PCA50_pred <- predict(tree_PCA50, newdata=PCA50_test, type="class")
class_rate_tree50 <- mean(tree_PCA50_pred==PCA50_test$Class)
class_rate_tree50 
```

The poor performance of these trees, and the large number of unimportant variables they both use, suggests that a lower number of PCs may be worth investigating.

The evident importance of PC13 suggests that we should repeat the process as such.

```{r}
PCA13_train <- data.frame(Class = as.factor(y_train), predict(pca_fit, X_train)[,1:13])
PCA13_test <- data.frame(Class = as.factor(y_test), predict(pca_fit, X_test)[,1:13])

# Full tree
tree_PCA13_full <- rpart(Class~., data=PCA13_train, method="class", minsplit=2, 
                         minbucket=1, cp=-1)
tree_PCA13_full

# Pruning:
printcp(tree_PCA13_full) 
# Minimum xerror: Tree 7, cp=0.011364
tree_PCA13 <- prune(tree_PCA13_full, cp=0.012)
rpart.plot(tree_PCA13, type=2, extra=4 )
tree_PCA13
```

This tree is pruned according to a different complexity parameter. All of the variables are of significant importance.

```{r}
tree_PCA13$variable.importance 

tree_PCA13_pred <- predict(tree_PCA13, newdata=PCA13_test, type="class")
class_rate_tree13 <- mean(tree_PCA13_pred==PCA13_test$Class)
class_rate_tree13

tab13 <- table(PCA13_test$Class, tree_PCA13_pred)
tab13
```

This tree returns a test accuracy of 70%, a slight improvement on the previous two trees. In the cross classification table, we also see that this tree produces significantly more false positives than false negatives, which seems preferrable in this context. There appears to be no benefit to including an extra 27 variables in a classification tree, so we proceed with 13 PCs.

Worth noting is that classification trees are subject to overfitting and high-variance. We turn to other tree based methods, contained within randomForest, to see whether the accuracy of our method can be improved.

We begin by applying bagging to a tree for data with 13 PCs.

```{r}
# Use randomForset, with mtry set to the number of variables
set.seed(101)
bag_PCA13 <- randomForest(Class~., data=PCA13_train, mtry=13, 
                          ntree=300) # 300 optimal
bag_PCA13 
```

Obtain an OOB estimate of error rate of 19%, and the cross corelation matrix already appears to be an improvement.

```{r}
varImpPlot(bag_PCA13)

bag_PCA13_pred <- predict(bag_PCA13, newdata=PCA13_test, type="class")
class_rate_bag <- mean(bag_PCA13_pred==PCA13_test$Class)
class_rate_bag
```

Whilst PC13 remains significantly the most important variable, the relative importance of other variables is different in this method than the previous tree.

The test accuracy is 82%, which is a vast improvement.

As a sanity check, we will verify that the benefit of using 13 PCs translates to this method by fitting a bagged model to data with 13 PCs.

```{r}
set.seed(101)
bag_PCA40 <- randomForest(Class~., data=PCA40_train, mtry=40, 
                          ntree=400) # 400 optimal
bag_PCA40 

varImpPlot(bag_PCA40)

bag_PCA40_pred <- predict(bag_PCA40, newdata=PCA40_test, type="class")
class_rate_bag2 <- mean(bag_PCA40_pred==PCA40_test$Class)
class_rate_bag2
```

This method has an OOB error rate of 21%. We see the expected pattern in the variable importance plot. The test accuracy is 80%, so we retain 13 PCs.

However, if we have observed anything so far it is that there is a dominant feature in this data, which leads to high correlation between bagged models, ultimately limiting the effect that bagging has on reducing variance.

We finally use a random forest method, which comabts the effects of dominant variables.

```{r}
# use randomForest without mtry argument
set.seed(101)
rf_PCA13 <- randomForest(Class~., data=PCA13_train, ntree=200) 
rf_PCA13 
varImpPlot(rf_PCA13)

rf_PCA13_pred <- predict(rf_PCA13, newdata=PCA13_test, type="class")
class_rate_rf <- mean(rf_PCA13_pred==PCA13_test$Class)
class_rate_rf 
```

This method has an OOB estimate of error rate of 18%. We see that all of the variables are less important. However, test accuracy is only 80%, so a decrease on the previous model.

```{r}
# use randomForest without mtry argument
set.seed(101)
rf_PCA40 <- randomForest(Class~., data=PCA40_train, ntree=200) 
rf_PCA40 
varImpPlot(rf_PCA40)

rf_PCA40_pred <- predict(rf_PCA40, newdata=PCA40_test, type="class")
class_rate_rf <- mean(rf_PCA40_pred==PCA40_test$Class)
class_rate_rf 
```

No, no magical change.

Of all of the classification tree methods, bagging produces the best results. And, across all of these models, using 13 PCs, which only capture 62% of variabiltiy, prodices best results. The classification tree of choice is obtained by applying bagging to data with 13 PCs, which achieves an accuracy of 82%. However, we allready know that this is not the most accurate classification method that has been tested for this data.
